---
title: "Simple univariate animal model"
linkTitle: "Univariate animal model"
weight: 4
description: >
  Fitting a simple univariate model in R.
---

{{% pageinfo %}}
Here we demonstrate how to fit the simplest univariate animal model to calculate the additive genetic variance component of a phenotypic variable, for example body size. In later sections we show how to build more complex models with fixed or random effects and how to compute heritability.
{{% /pageinfo %}}


 is the simplest of all animal models and assumes no repeated measures across years or individuals.


# Example data

For this tutorial we will use the simulated gryphon dataset ([download zip file](data/gryphons.zip)).

```{r}
phenotypicdata <- read.csv("data/gryphon.csv")
pedigreedata <- read.csv("data/gryphonped.csv")
```


# The univariate animal model

**Definitions**

Sample code in the following examples includes the following variables:

* Response variable: Size
* Fixed effects: Intercept ($\mu$)
* Random effects: Additive genetic variance
* Data containing phenotypic information: phenotypicdata
* Data containing pedigree data:pedigreedata

We first fit the simplest possible animal model: no fixed effects apart from the interecept, a single random effect (the breeding values, associated with the additive genetic variance), and Gaussian redisuals.

{{% pageinfo %}}
Here it is not too difficult to describe the model in plain English, but for more complex models a mathematical equation may prove easier, much shorter and less ambiguous. Here we could write the model as $y_i = \mu + a_i + \epsilon_i$, where $y_i $ is the response for individual $i$, with the residuals assumed to follow a normal distribution of mean zero and variance $\sigma_R^2$, which we can write as $\mathbf{\epsilon} \sim N(0,\mathbf{I}\sigma_R^2)$, where $\mathbf{I}$ is an identity matrix; and with the breeding values $\mathbf{a} \sim N(0,\mathbf{A} \sigma_A^2$, where $\mathbf{A}$ is the pairwise additive genetic relatedness matrix.

{{% pageinfo %}}



## Using MCMCglmm

Here is the simplest implementation of an animal model in MCMCglmm:

```{r, cache=TRUE, message=FALSE, results='hide'}
library(MCMCglmm) #load the package

inverseAmatrix <- inverseA(pedigree = pedigreedata)$Ainv # compute the inverse relatedness matrix

model1.1<-MCMCglmm(birth_weight~1, #Response and Fixed effect formula
                   random=~id, # Random effect formula
          ginverse = list(id=inverseAmatrix), # correlations among random effect levels (here breeding values)
          data=phenotypicdata)# data set
```

Note the use of the argument `ginverse`


```{r}
summary(model1.1)
```


## Using ASReml

From R:
```r
model1<-asreml(fixed=SIZE~ 1                    #1  
  , random= ~ped(ANIMAL,var=T,init=1)           #2  
  , data=phenotypicdata                         #3
  ,ginverse=list(ANIMAL=pedigreedata_inverse)   #4
  , na.method.X="omit", na.method.Y="omit")     #5
```






 
