---
title: "Univariate animal model using brms"
linkTitle: "brms"
weight: 6
description: >
  Univariate animal model using brms
output: 
  html_document: 
    keep_md: yes
---


# Example data

We will use the simulated gryphon dataset ( [download zip file](/docs/data/gryphons.zip) ).

We need to load both the phenotypic data `gryphon.csv` and the pedigree `gryphonped.csv`.

```{r}
phenotypicdata <- read.csv("data/gryphon.csv")
pedigreedata <- read.csv("data/gryphonped.csv")
```

The phenotypic data look like this:

```{r}
head(phenotypicdata)
```

We will use `birth_weight` as a response variable.

And the pedigree looks like this:

```{r}
head(pedigreedata)
tail(pedigreedata)
```


# Simplest animal model

Here is the simplest implementation of an animal model in brms.

First, we load the package:

```{r, message=FALSE, results='hide'}
library(brms);library(nadiv)
```

To be able to fit an animal model, brms needs the relatedness (relationship) matrix of the pedigree and not its inverse (as in other softwares). This can be estimated using the nadiv package created by Pr. Matthew Wolak (https://cran.r-project.org/web/packages/nadiv/index.html).

```{r, cache=TRUE}
Amat <- as.matrix(nadiv::makeA(pedigreedata))
```

Now we can fit the model of `birth_weight` to estimate three parameters:

* an additive genetic variance (corresponding to the `id` column) 
* a residual variance
* an intercept

```{r, echo=FALSE}
set.seed(12345)
```


```{r, cache=TRUE, message=FALSE, results='hide',warning=F}
brms_m1.1 <- brm(
  birth_weight ~ 1 + #Response and Fixed effect formula
    (1 | gr(id, cov = Amat)),# Random effect formula & correlations among random effect levels (here breeding values)
  data = phenotypicdata,# data set
  data2 = list(Amat = Amat), # Amatrix
  family = gaussian(), # family
  chains = 2, cores = 2, iter = 2000 # two mcmc chains run on two cores (in parallel) for 2000 iterations each
)
```


```{r brmplot1-plot}
plot(brms_m1.1)
```




