---
title: "Animal model with GRM"
linkTitle: "Heritability of body condition in the Seychelles warbler"
weight: 4
math: true
description: >
  This is a tutorial about how to fit an 'animal model' with GRM (genetic relationship matrix). We provide the sample with the Seychelles warbler data subset including body mass, tarsus length, and also other factors data of 36 individuals and the GRM were calculated from 10k snps genomic data filtered with PLINK -maf0.05 ).
output: 
  html_document: 
    keep_md: yes
---

```{r setup_cache, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

> This is a tutorial about how to fit an 'animal model' with GRM (genetic relationship matrix). We provide the sample with the Seychelles warbler data subset including body mass, tarsus length, and also other factors data of 36 individuals and the GRM were calculated from 10k snps genomic data filtered with PLINK -maf0.05 ).

[dataset](/docs/Univariate/GRM/dataset_bodycondition_sw.csv) and GRM files can be downloaded from the github

# Why do you want a GRM?

- NGS is (relatively) getting cheaper
- More information to estimate the relatedness, sometimes (like the warblers) the social pedigree can contain errors due to wrong observations or extra-pair paternity...
- No pedigree information in your study
- ...

> Before go to create a GRM, it is better to check the population structure by simply checking with a PCA plot.

# The method to generate GRM

There are many methods and tools (e.g., PLINK, NgsRelate\* or R packages like 'qgg') to estimate the relatedness between individuals and generate a GRM, here is an example by using GCTA -grm flag. Note: NgsRelate accepts genotype likelihood input and vcf input

```{r}
#system("gcta64 or gcta64.exe --bfile filterd.final.autosome.maf0.05_bodymass_h2_gwas_50ind_10k_WAMBAM  --make-grm --autosome-num 29 --out filterd.final.autosome.maf0.05.36ind.10k.WAMBAM")
```

There are 3 output files of GCTA: grm.bin, grm.id, grm.N.bin, you need to put them under the same directory

Here, we fit the data into animal model in brms

First, you need to install and load all the packages, you can check brms usage tutorial to know how to setup first

```{r}
#install.packages(c("readr","genio","brms","cmdstanr","rstan"))
##setup your work directory
setwd("~/wamwiki.github.io/content/en/docs/Univariate/GRM/")
library(readr)
library(genio)
library(MASS)
library(brms)
library(cmdstanr)
library(tidybayes)
library(rstan)
library(dplyr)
library(ggplot2)
```

load the dataset, GRM

```{r}
dataset <- read.csv("dataset_bodycondition_sw.csv")
grm.data <- read_grm("filterd.final.autosome.maf0.05.36ind.10k.WAMBAM.grm")
GRM_pd <- as.matrix(Matrix::nearPD(grm.data$kinship)$mat)
```

Check if the dataset and GRM with the same individual

```{r}
dataset_birds <- unique(dataset$BirdID)
cat("Birds in dataset:", length(dataset_birds), "\n")
cat("Birds in GRM:", nrow(GRM_pd), "\n")
cat("Overlap:", sum(dataset_birds %in% rownames(GRM_pd)), "\n")
cat("Missing from GRM:", sum(!dataset_birds %in% rownames(GRM_pd)), "\n")
cat("Missing from model:", sum(!rownames(GRM_pd) %in% dataset_birds), "\n")
```

Keep only birds present in model and GRM

```{r}
birds_in_both <- dataset_birds[dataset_birds %in% rownames(GRM_pd)]
cat("Birds to use in GRM animal model:", length(birds_in_both), "\n")

GRM_sub <- GRM_pd[as.character(birds_in_both), as.character(birds_in_both)]

dim(GRM_sub)

data_final <- dataset[dataset$BirdID %in% birds_in_both, ]
```

Be careful with the prior, but in general the default prior works well. Here we used the default prior get from get_prior() function.

```{r}
# get priors from get_prior() function
my_priors <- c(prior(normal(0, 1),class = b,coef = age_year),
               prior(normal(0, 0.5),class = b,coef = Iage_yearE2), 
               prior(normal(0, 2),class = b,coef = RightTarsus), 
               prior(normal(0, 1),class = b,coef = SexEstimate),  
               prior(normal(0, 1),class = b,coef = summer),
               prior(normal(0, 1),class = b,coef = minutes_s),
               prior(normal(0, 1),class = b,coef = avg_invert),
               prior(normal(0, 1),class = b,coef = group_size),
               # fixed effects priors were provided above
               prior(normal(15.5, 3),class = Intercept), # intercept prior
               prior(student_t(3, 0, 1), class = sd), # random effect sd priors
               prior(student_t(3, 0, 1), class = sigma)) # residual sd prior
```

We set chains = 4 to let the model to run 4 MCMC chains at the same time, and also saved as rda file, brms model structure:

```{r model}
model_grm <- brm(
  BodyMass ~ age_year + I(age_year^2) + RightTarsus + SexEstimate + summer +minutes_s + avg_invert + group_size + (1 | gr(animal, cov = GRM_pd)) + (1 | BirdID) + (1 | BirthYear) + (1 | CatchYear) + (1 | Observer),
  data    = data_final,
  data2   = list(GRM_pd = GRM_pd),
  family  = gaussian(),
  prior   = my_priors,
  warmup  = 1500,
  iter    = 11500,
  chains  = 4,
  cores   = 4,
  seed    = 123,
  file    = "m_brms_GRM",
  backend = "cmdstanr")
```

check the ESS, Rhat and check_hmc_diagnostics() to evaluate the animal model:

```{R}
summary(model_grm)
rstan::check_hmc_diagnostics(model_grm$fit)
```

Also check the mcmc traces if they mixed well:

```{r figure1, fig.width=8, fig.height=6, out.width="80%", fig.path="figures/"}
plot(model_grm)
```

The heritability (h2) can be computed:

```{r}
#Vf (fixed effect variance) calculation
mu_fixed <- posterior_linpred(model_grm, re_formula = NA)
# variance per posterior draw
Vf_GRM <- apply(mu_fixed, 1, var)

h2_draws_Vf_GRM <- model_grm %>%
  spread_draws(sd_animal__Intercept, sd_BirdID__Intercept, sd_BirthYear__Intercept,
               sd_CatchYear__Intercept, sd_Observer__Intercept, sigma) %>%
  mutate(Vf  = Vf_GRM,
         Vg  = sd_animal__Intercept^2,
         Vpe = sd_BirdID__Intercept^2,
         Vby = sd_BirthYear__Intercept^2,
         Vcy = sd_CatchYear__Intercept^2,
         Vobs= sd_Observer__Intercept^2,
         Ve  = sigma^2,
         Vp  = Vg + Vpe + Vby + Vcy + Vobs + Ve + Vf,
         h2  = Vg / Vp,
         repeatability = (Vg + Vpe) / Vp,
         CVa = 100 * (sd_animal__Intercept / mean(data_final$BodyMass, na.rm = TRUE)))
```

We can also get the point estimation:

```{r}
h2_draws_Vf_GRM %>% mean_hdi(h2)
h2_draws_Vf_GRM %>% mean_hdi(repeatability)
h2_draws_Vf_GRM %>% mean_hdi(CVa)
```

Plot the h2

```{r figure2,fig.width=8, fig.height=6, out.width="80%", fig.path="figures/"}
ggplot(h2_draws_Vf_GRM, aes(x = h2)) +
  stat_halfeye(.width = c(0.95), fill = "grey70", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4) +
  scale_x_continuous(limits = c(0, NA)) +
  labs(x = expression(h^2), y = NULL) +
  theme_classic(base_size = 16)
```