---
title: "Simulating genetic variation in a trait"
linkTitle: "Simulating Traits"
weight: 5
description: >
  Simulating genetic variation in a trait 
output: 
  html_document: 
    keep_md: yes
---

# Simulating breeding values

Simulating data can be useful for many reasons, such as testing models, conducting power analysis, post-hoc model validation, etc. Data simulation involves (pseudo-)random number generation, and so the results of simulations will vary each time a simulation is run (like an MCMC model). We therefore do not expect to get back the exact parameters we used to simulate, but we do expect that over many simulations, the mean parameters values form the simulation should converge on the true, simulated values. 

To simulate breeding values (or additive genetic effects), we first need a pedigree to simulate values from. We will use the Gryphon pedigree, which is in the `pedtricks` package.

```{r, message=FALSE}
library(pedtricks)
data(gryphons)
ped <- fix_ped(gryphons[,1:3])
head(ped)
```

The simplest way to simulate breeding values is using the `rbv()` function in the `MCMCglmm` package. 

```{r, message=FALSE}
library(MCMCglmm)
bv <- rbv(ped, G=0.3)
```

Here `G=...` refers to the G matrix. In our univariate case, G just represents the variance, which we have specified as 0.3.

This in itself may not be that useful, so we can consider how to use this to build up a trait.


# Simulating a trait

If we consider that $P=G+E$, then we can make a simple trait by adding our simulated breeding values above to some simulated residual values. To do this, we can use the `rnorm()` function. Importantly, `rnorm()` takes the standard deviation as an arguments, not the variance, so a little care need to be taken thinking about specifying this compared with `rbv()`.

If we aim for a total variance of one, this makes it easy to simulate a certain heritability. As $V_P = V_A + V_E$, if we aim for a phenotypic variance of 1, then given the variance in breeding values from above (0.3), we need our residuals to have a variance of 1-0.3 = 0.7. This will give a heritability ($\frac{V_A}{V_P}$) of 0.3.


```{r}
e <- rnorm(nrow(ped), mean=0, sd = sqrt(0.7))
p <- bv + e
```

We can test that this works by running an animal model. Here we use gremlin, but see [https://wildanimalmodels.org/docs/univariate/](https://wildanimalmodels.org/docs/univariate/) for how to run this model with other packages. 

```{r, message=FALSE}
library(gremlin)
library(nadiv)

data <- data.frame(p=p, animal=ped[,1])
Ainv <- makeAinv(ped)$Ainv
mod <- gremlin(p~1, random=~ animal,data=data,ginverse=list(animal=Ainv))
summary(mod)
```

As stated above, simulations have a stochastic component, so we do not expect the variance estimates to be exactly the same as what we simulated ($V_A=0.3$ and $V_E=0.7$). However, they are very close, so we can be fair happy that our simulation is doing a good job. 


# Adding complexity
Whilst basic simulations such as this are fairly easy to construct, building more complexity into simulations can be challenging. The [`squidSim` R package](https://github.com/squidgroup/squidSim) facilitates building in added complexity, such as repeated measures, maternal effects (environmental and genetic), multivariate phenotypes, sex-specific variances and GxE. A detailed vignette for the `squidSim` package (including how to simulate additive genetic effects) can be found [here](https://squidgroup.org/squidSim_vignette/index.html). 

To demonstration a more complex example, we can add in maternal and cohort effects.

First, we need to create a data structure that contains the hierarchical structure we need to simulate the other effects. To do this we can subset the gryphon data, as not all individuals have a known mother, and then take the id, dam and cohort columns that have the hierarchical structure we need.

```{r, message=FALSE}
library(squidSim)

ds<- subset(gryphons, !is.na(dam))[,c("id","dam","cohort")]
head(ds)
```

We then input this data structure into the `data_structure=...` argument, and use the `pedigree = ...` argument to link the pedigree to the id in the data structure. The `parameters = ...` argument then allows us to list the respective variances (`vcov=...`) for each component.

```{r}
squid_data <- simulate_population(
  data_structure = ds, 
 
  pedigree = list(id=ped),
  
  parameters =list(
    id = list(
      vcov = 0.3
    ),
    dam = list(
      vcov = 0.1
    ),
    cohort = list(
      vcov = 0.2
    ),
    residual = list(
      vcov = 0.4
    )
  )
)

```

To extract the simulate data we run 
```{r}
data2 <- get_population_data(squid_data)
head(data2)
```
in here we have the trait values (`y`), as well as the data structure and all the simulated effects. 

We can then run an animal model on this 

```{r}
Ainv <- makeAinv(ped)$Ainv
data2$id<-as.factor(data2$id)
data2$dam<-as.factor(data2$dam)
data2$cohort<-as.factor(data2$cohort)
mod2 <- gremlin(y~1, random=~ id + dam + cohort,data=data2,ginverse=list(id=Ainv))
summary(mod2)
```

Here again we do not return the exact values that we simulated, but we can see that they are similar. 
